---
title: "IRT_analysis"
author: "Tuomas Aivelo"
date: "22 5 2020"
output: html_document
---
This is a commented R script for the analysis for the manuscript "Factors explaining students’ attitudes towards learning in genetics and belief in genetic determinism" by Tuomas Aivelo and Anna Uitto

First we install all the needed packages:

```{r}
deps = c("car","mirt","likert","dplyr","forcats","dotwhisker","tidyr","tidyverse", "rfigshare");

for (dep in deps){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}

```

The data is downloaded from the Figshare repository and processed to the needed format

```{r}
url <- "https://ndownloader.figshare.com/files/22772264"
download.file(url, destfile="data.csv", 'libcurl')

data <- read.csv(file="data.csv", sep=";")
rownames(data) <-data$form
data$course_tot <- as.numeric(data$course_tot)
data$gender <- as.factor(data$gender)
data$age <- as.numeric(data$age)
data$textbook <- as.factor(data$textbook)
data$teacher_app <- as.factor(data$teacher_app)

```

First we format the data so we can do Figure 1. We take the subset of Likert scale answers for items 1 to 25 and their statements. These are inputs to the Figure made with likert package.

```{r}

# raw answers for Figure 1
lik <- data[,15:39]
lik[lik==1] <- "Strongly agree"
lik[lik==2] <- "Somewhat agree"
lik[lik==3] <- "Not disagree or agree"
lik[lik==4] <- "Somewhat disagree"
lik[lik==5] <-"Strongly disagree"

lik[]<- lapply(lik, as.factor)

#Figure 1 

factlevel <- c("Strongly disagree", "Somewhat disagree","Not disagree or agree", "Somewhat agree","Strongly agree")

factfunc <- function(mydata, factlevel){
  factor(mydata, 
         levels=factlevel, 
         ordered = TRUE)
  fct_unify(mydata, 
            levels=factlevel) 
}

lik_ordr <- factfunc(lik, factlevel) %>%as.data.frame()

names(lik_ordr) <- c(
  X1="Genetics is a difficult field of biology.",
  X2="Modification of human genomes is ethically acceptable if it allows for curing genetic disorders. ",
  X3="Genes determine all the traits in humans.",
  X4="Genetically modified alimentary products are useful for humankind.",
  X5="It is boring to study genetics.",
  X6="It is impossible for me to achieve good results in learning genetics.",
  X7="I need knowledge on genetics in my future studies.",
  X8="If the genome of a person is known, we know exactly which diseases they will have.",
  X9="I think I excel in genetics.",
  X10="Stem cell research done on human embryo should be illegal.",
  X11="Human intelligence is determined by their genes.",
  X12="I do not need in future what I have studied in genetics course so far.",
  X13="Usually we have interesting exercises in genetics.",
  X14="Genetics literacy is required in everyday life.",
  X15="Many things in genetics are difficult.",
  X16="I like to study genetics.",
  X17="Studying human genomes leads to parents selecting offspring who is genetically favorable.",
  X18="I can solve even difficult genetics exercises.",
  X19="I think knowing genetics is important.",
  X20="The effects of genes in human happiness is marginal.",
  X21="Genetics is one of my favorite topics in biology",
  X22="It is acceptable that scientists develop deadlier pathogens in laboratory through genetic modification if that allows for new cures to diseases.",
  X23="Understanding genetics is more and more important in future.",
  X24="Some people are more succesful in school due to their genes.",
  X25="I like biology lessons where we discuss genetics.")

order <- c("Genetics is a difficult field of biology.","It is impossible for me to achieve good results in learning genetics.","I think I excel in genetics.","Many things in genetics are difficult.","I can solve even difficult genetics exercises.",
           "It is boring to study genetics.","Usually we have interesting exercises in genetics.","I like to study genetics.","Genetics is one of my favorite topics in biology","I like biology lessons where we discuss genetics.",
           "I need knowledge on genetics in my future studies.","I do not need in future what I have studied in genetics course so far.","Genetics literacy is required in everyday life.","I think knowing genetics is important.","Understanding genetics is more and more important in future.",
           "Modification of human genomes is ethically acceptable if it allows for curing genetic disorders. ","Genetically modified alimentary products are useful for humankind.","Stem cell research done on human embryo should be illegal.","Studying human genomes leads to parents selecting offspring who is genetically favorable.","It is acceptable that scientists develop deadlier pathogens in laboratory through genetic modification if that allows for new cures to diseases.",
           "Genes determine all the traits in humans.","If the genome of a person is known, we know exactly which diseases they will have.","Human intelligence is determined by their genes.","The effects of genes in human happiness is marginal.","Some people are more succesful in school due to their genes.")

liklik <- likert(lik_ordr)

plot(liklik, ordered=FALSE, group.order=order)

```

Then we go to the actual analysis. First we need to reverse items 1,5,6,10,12,15,17 and 20.

```{r}

data$X1<-recode(data$X1,'1'=5L,'2'=4L,'4'=2L,'5'=1L)
data$X5<-recode(data$X5,'1'=5L,'2'=4L,'4'=2L,'5'=1L)
data$X6<-recode(data$X6,'1'=5L,'2'=4L,'4'=2L,'5'=1L)
data$X10<-recode(data$X10,'1'=5L,'2'=4L,'4'=2L,'5'=1L)
data$X12<-recode(data$X12,'1'=5L,'2'=4L,'4'=2L,'5'=1L)
data$X15<-recode(data$X15,'1'=5L,'2'=4L,'4'=2L,'5'=1L)
data$X17<-recode(data$X17,'1'=5L,'2'=4L,'4'=2L,'5'=1L)
data$X20<-recode(data$X20,'1'=5L,'2'=4L,'4'=2L,'5'=1L)

```

Then we look at the descriptive statistics. 

```{r}
summary(data)

```

Exploratory analysis uses all of the times and every respondent.Summary is the output which is given in Table S1 of the manuscript. 

```{r}

likert <- data[,15:39]
expl5 <- mirt(likert, 5, itemtype="gpcm", method="MHRM")
summary(expl5) 
```

Then we define the model for the confirmatory analysis based on our items. We remove poorly performing items based on the previous step. Used dimesions are F1: liking genetics, F2: attitudes towards gene technology, F3: belief in genetic determinism, F4: perceived utility of genetics, F5: selfconcept in genetics

```{r}


items <- likert[,c(1:5,7:13,15,16,18:19,21:22,25)]
model <- mirt.model('
                    F1 = 5,12,14,17,19
                    F2 = 2,4,9,18
                    F3 = 3,7,10
                    F4 = 6,11,16
                    F5 = 1,8,13,15
                    COV = F1*F2*F3*F4*F5
                      ')


```

We also define a model to calculate the priors for the model run to facilitate in reaching the sensible model. 
```{r}
model_val <- mirt.model('
                    F1 = 5,12,14,17,19
                    F2 = 2,4,9,18
                    F3 = 3,7,10
                    F4 = 6,11,16
                    F5 = 1,8,13,15
                      ')
values <- mirt(items, model_val, pars = 'values', method = "MHRM", itemtype = 'gpcm')
```

Then we run the confirmatory model. This gives us the model fit statistics, item fit statistics, Q3 statistics, person-fit statistics. Based on the person-fit, we remove part of the respondents as outliers.

```{r}

mod_gpcm <- mirt(items, model, pars=values, itemtype="gpcm", method="MHRM")
M2(mod_gpcm, type="M2*", na.rm=TRUE, QMC=TRUE) 

fit<-itemfit(mod_gpcm, na.rm=TRUE, QMC=TRUE)
p.adjust(fit$p.S_X2, method = 'fdr')

resid <- residuals(mod_gpcm, type = 'Q3')
sum(resid>0.2) 

mod_gpcmt <- mirt(na.omit(likert), model, itemtype="gpcm", method="MHRM")
Theta <- fscores(mod_gpcmt, method = 'MAP', QMC=TRUE)

persfit <- personfit(mod_gpcmt, Theta=Theta, method ='MAP', QMC=TRUE)
misfits <- subset(persfit, Zh< -2)

likert2 <- likert[-as.numeric(rownames(misfits)),]
```

We refine the data for the explanatory model, including scaling of the variables

```{r}

data2 <- data[-as.numeric(rownames(misfits)),]
cov <- data2[,c(2,10:14)]
cov$school <- as.numeric(cov$school)
cov$age <- scale(cov$age)
cov$course_tot <- scale(cov$course_tot)
cov <- data.frame(cov)

```

Then we run the confirmatory model. Coefficients are used for the Table 2 of the manuscript on the correlations between factors.

```{r}
bdmod<- mixedmirt(likert2, cov, model, itemtype="gpcm", 
                  fixed = ~0, lr.random = ~ 1|school,
                  lr.fixed = ~ gender + course_tot + textbook + teacher_app,
                  technical = list(removeEmptyRows = TRUE))


coef(bdmod)
```

Finally, we use the coefficients of the model to produce Figure 2 of the manuscript.

```{r}
coef <- coef(bdmod)$lr.betas
coef <- t(coef)
comb <- rownames(coef)

comb <- recode(comb, F1_teacher_app2 = "F1_teacherapp2", F2_teacher_app2 = "F2_teacherapp2", F3_teacher_app2 = "F3_teacherapp2", F4_teacher_app2 = "F4_teacherapp2", F5_teacher_app2 = "F5_teacherapp2",
       F1_teacher_app3 = "F1_teacherapp3", F2_teacher_app3 = "F2_teacherapp3", F3_teacher_app3 = "F3_teacherapp3", F4_teacher_app3 = "F4_teacherapp3", F5_teacher_app3 = "F5_teacherapp3")

modl <- strsplit(comb,"_")
model <- (sapply(modl, "[",1))
submodel <- sapply(modl, "[",2)
term <- recode(submodel, teacherapp2 = "teacherapp",  teacherapp3 = "teacherapp")

coef <- cbind(coef, model, submodel,term)
coef <- coef[c(2:6,8:12,14:18,20:24,26:30),]

work <- as_tibble(coef)

work <- work %>% 
  rename(
    estimate = par
  )

work <- work %>% 
  mutate_at(vars(estimate, CI_2.5, CI_97.5), as.numeric)


ggplot(work, aes(x = estimate, xmin = CI_2.5, xmax = CI_97.5, y = as.factor(term), colour = submodel))+
  geom_point()+
  geom_errorbar(width=0.1)+
  theme_bw()+
  facet_grid(cols=vars(model))
```

